Spatial variation of the native colon microbiota in healthy adults

### 2018-01-30 submission/manuscript.tex
took figures out of manuscript.Rmd'
reviewer revisions completed

The human colon is an ecosystem comprised of numerous microenvironments
that select for different microbiota. Concentrations of oxygen, water,
and anti-microbial peptides change along the gut axis and influence
which microbiota reside in each location. Microenvironments differ not
only longitudinally along the colon, but also radially from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. The identity of these
specific microbiota and communities are important for understanding the
etiology of complex diseases such as Colorectal Cancer (CRC) and
Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
accelerated by perturbations the structure of the gut microbiota (1--3).
The manifestations of these diseases are known to vary based upon the
location in which they occur. For instance, CRC that arises in the
distal (left) colon are of hindgut origin and tend to have large
chromosomal alterations indicative of chromosomal instability (1). In
contrast, CRC arising in the proximal (right) colon tumors are of midgut
origin and tend to be sessile and microsatellite instable (MSI with BRAF
and KRAS mutations) (1). In addition to the environmental gradients
within the colon, the distal and proximal sides of the colon differ in
the amount of inflammation present and the genomic instability of
precancerous cells, respectively (1,4,5). In IBD patients, disease
occurring in the distal colon extending proximally is usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur
anywhere along the GI tract, most commonly in the ileum and the cecum
(2). UC presents as continuous disease with only mucosal involvement,
where as CD has skip lesions and full thickness involvement that may
cause abscesses, strictures and fistulas (2). Thus, given the varied
physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbiota and
their metabolites likely vary as well, and may influence the
heterogeneous disease prognoses of IBD and CRC. Because CRC can be a
long-term complication of IBD, the distribution of microbiota is
important to understanding the pathophysiology of both diseases.

@@

The few existing profiles of the microbial spatial variation of the
colon have been limited by sample collection methods. The majority of
human gut microbiome studies have been performed on whole shed feces or
on samples collected during colonoscopy or surgery (5). While the latter
method allows investigators to acquire samples from inside the human
colon, typically these procedures are preceded by the use of bowel
preparation methods such as the consumption of laxatives to cleanse the
bowel. Bowel preparation is essential for detecting cancerous or
precancerous lesions in the colon, but complicates microbiome profiling
as the chemicals strip the bowel of contents and disrupt the mucosal
layer (13,14). As such, what little information we do have about the
spatial distribution of the microbiota in the proximal and distal colon
is confounded by the bowel preparation procedure.

@@

The procedures in this study and consent were approved by the
Institutional Review Board at the University of Michigan Health System
with protocol number HUM00082721. Subjects were recruited using the
online recruitment platform and were pre-screened prior to enrollment in
the study. Exclusion criteria included: use of asprin or NSAIDs within 7
days, use of antibiotics within 3 months, current use of anticoagulants,
known allergies to Fentanyl, Versed and Benadryl, prior history of colon
disease, diabetes, abdominal surgery, respiratory, liver, kidney or
brain impairments, undergoing current chemotherapy or radiation
treatment and subjects that were pregnant or trying to conceive. 20
subjects that met the criteria were selected and provided signed
informed consent prior to the procedure. There were 13 female and 7 male
subjects ranging in age from 25 to 64.

@@

At a baseline visit, subjects gave consent and were given a home
collection stool kit (Zymo). At least one week prior to the scheduled
colonoscopy, subjects collected whole stool at home and shipped the
samples to a research coordinator on ice. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On the procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Subjects were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and jumbo
biopsy forceps used to collect the luminal contents. Two luminal samples
were collected and the contents immediately deposited into RNAlater
(Fisher) and flash-frozen in liquid nitrogen. The forceps were withdrawn
and new biopsy forceps were used to collect mucosal biopsies on sections
of the colon that were pink and free of stool matter. Three mucosal
biopsies were collected and flash-frozen in RNAlater. These samples
comprised the distal colon samples. The sigmoidoscope was then withdrawn
and a pediatric colonoscope was inserted to reach the proximal colon.
Samples were then collected in the same manner as was done in the distal
colon and the colonoscope withdrawn. All samples were stored at
-80\(^\circ\)C.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described (15). Sequences were curated
using the mothur software as described previously (16). The sequences
were assigned a taxonomic classification using a naive Bayesian
classifier trained using a 16S rRNA gene training set from the Ribosomal
Database Project (RDP) (17) and clustered into operational taxonomic
units (OTUs) based on a 97\% similarity cutoff. Sequencing and analysis
of a mock community revealed the error rate to be 0.018\%. Samples were
rarefied to 4231 sequences per sample in order to reduce the effects of
uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
1.39.5 (16). \(\theta\)\textsubscript{YC} distances were calculated to
determine the dissimilarity between two samples. Random Forest
classification models were built using the AUCRF R package using a
leave-one-subject out approach (18). For each model the data were split
into a 19-subject training set and a 1-subject test set. The model was
built and cross-validated using AUCRFcv on the training set. The model
was then tested on the left-out patient. This process was repeated
iteratively for all subjects and results plotted as Reciever Operator
Characteristic curves using the pROC R package (19). Resultant models
were used to identify the OTUs that were most important for classifying
each location. Species-level information for sequences of interest was
obtained by aligning the sequences to the GenBank nucleotide databse
using blastn. The species name was only used if the identity score was
\(\ge\) 99\% over the full-length of the contig and matched a single
reference.

@@

When examining the relative abundance of the dominant phyla at each site
(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
of variation. This likely reflects not only the variability between
human subjects, casued by differences in age, sex, diet, but also
spatial ``patchiness'' in the gut microbiome. One study noted that the
bacteria recoverable from the same mucosal sample location can be vastly
different when the samples are taken just 1 cm away from each other
(26). Similar patchiness was also observed in luminal contents and fecal
samples themselves; there was separation of different interacting
microbes along the length of a stool sample, for instance (27). That
said, across our samples, the mucosal samples harbor more
\emph{Proteobacteria}, consistent with previous studies comparing
mucosal swabs to luminal content in humans (4). Hence, the conclusions
we were able to draw from phyla analysis may have been impacted by
inter-subject patchiness.

@@

In addition to identifying features that are specific to each side of
the gut, the ability of the Random Forest to classify samples can serve
as a proxy for similarity. That is, a higher AUC value indicates the
samples are more efficiently classfied (and thus more different) than a
model with a lower AUC value. For instance, the model separating the
proximal and distal mucosa had an AUC of 0.850 whereas the model for
classifying the proximal and distal lumen had a much lower AUC of 0.580.
Further, the latter model required 44 OTUs to best separate the samples
whereas the models separating the mucosa only needed 10 OTUs. The much
lower AUC and need for a high number of features compared to other
models suggest these locations are the most similar of the comparisons
tested. We speculate that the model was less effective at classifying
the proximal and distal luminal contents because the mucosal
microenvironments have variable selective pressure along the colon than
the luminal microenvironments.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (9) and have oncogenic
properties in cell culture and in mice (28). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the distal mucosa. Not much is
known about the distribution of \emph{P. asacharolytica} along the
colon, but given its anaerobic lifestyle and asacharolytic metabolism,
it is not surprising that it resides in the less-oygen-rich and
protein-rich distal mucosa (4). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} was more commonly detected
on proximal-sided tumors, and distribution of \emph{F. nucleatum}
decreased along the colon to rectum (29). In another study,
\emph{Fusobacterium} was associated with MSI with BRAF and KRAS
mutations, molecular features of proximal CRC (30). Of the 8 (40\%)
individuals positive for \emph{F. nucleatum} in the present study, the
bacterium was spread across the proximal mucosa, distal lumen and distal
mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors
suggests that \emph{Fusobacteria} species are more commonly found on
proximal tumors and in biofilms, indicating that it is not only the
presence of the bacterium but the structure of the tumor community that
contributes to \emph{Fusobacterium's} role in tumorigenesis (7).
Finally, \emph{Fusobacterium} and \emph{Porphyromonas} populations not
only co-occur on CRC tumors but also to synergize to promote
tumorigenesis in an oral cancer model (31,32). Further analysis of the
distribution and activities of these pathogens along the colon is needed
to elucidate a mechanism for development of CRC or IBD subtypes in the
proximal or distal colon.

The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD (25).
Laboratory experiments with these isolates have shown that
disease-isolated \emph{F. nucleatum} are more invasive and stimulate
more TNF-\(\alpha\) production than strains from healthy individuals
(10), suggesting the bacteria may increase inflammation in the gut as
well (33). \emph{F. varium} isolated from UC patients caused colonic
ulcers in an experimental mouse model (34). \emph{F. varium} was only
detected in three subjects and two of those samples were isolated form
the proximal mucosa (Supplementary Fig. S1B). \emph{F. varium} is most
commonly isolated from UC patient biopsies from the ileum or cecum (35),
suggesting this species may exhibit preference for the different
environmental conditions of these gastrointestinal sites. Further work
will assess how gut environment may select for species which may then
cause localized disease.

@@

By revealing specific differences in microbial populations at each
location in the gut via sampling an unprepared bowel, we can begin to
form hypotheses about how specific host-microbe interactions can affect
disease progression of proximal and distal CRC and IBD subtypes. A
better understanding of microbial activities in the gut can enhance
microbiome-based screening and treatment modalities for these colon
diseases.

@@

\hypertarget{ref-Hong2011}{}
26. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
Pyrosequencing-based analysis of the mucosal microbiota in healthy
individuals reveals ubiquitous bacterial groups and micro-heterogeneity.
Ahmed N, editor. PLoS ONE {[}Internet{]}. Public Library of Science
(PLoS); 2011;6:e25042. Available from:
\url{https://doi.org/10.1371/journal.pone.0025042}

\hypertarget{ref-Stearns2011}{}
27. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
tract. Scientific Reports {[}Internet{]}. Springer Nature; 2011;1.
Available from: \url{https://doi.org/10.1038/srep00170}

\hypertarget{ref-Sears2014}{}
28. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
from: \url{https://doi.org/10.1016/j.chom.2014.02.007}

\hypertarget{ref-Mima2016}{}
29. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
Fusobacterium nucleatum in colorectal carcinoma tissue according to
tumor location. Clinical and Translational Gastroenterology
{[}Internet{]}. Springer Nature; 2016;7:e200. Available from:
\url{https://doi.org/10.1038/ctg.2016.53}

\hypertarget{ref-Tahara_2014}{}
30. Tahara T, Yamamoto E, Suzuki H, Maruyama R, Chung W, Garriga J, et
al. Fusobacterium in colonic flora and molecular features of colorectal
carcinoma. Cancer Research {[}Internet{]}. American Association for
Cancer Research (AACR); 2014;74:1311--8. Available from:
\url{https://doi.org/10.1158\%2F0008-5472.can-13-1865}

\hypertarget{ref-Whitmore2014}{}
31. Whitmore SE, Lamont RJ. Oral bacteria and cancer. Goldman WE,
editor. PLoS Pathogens {[}Internet{]}. Public Library of Science (PLoS);
2014;10:e1003933. Available from:
\url{https://doi.org/10.1371/journal.ppat.1003933}

\hypertarget{ref-Flynn2016}{}
32. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
oral bacteria in colorectal cancer. McMahon K, editor. mSphere
{[}Internet{]}. American Society for Microbiology; 2016;1:e00102--16.
Available from: \url{https://doi.org/10.1128/msphere.00102-16}

\hypertarget{ref-Dharmani2011}{}
33. Dharmani P, Strauss J, Ambrose C, Allen-Vercoe E, Chadee K.
Fusobacterium nucleatum infection of colonic cells stimulates MUC2 mucin
and tumor necrosis factor alpha. Infection and Immunity {[}Internet{]}.
American Society for Microbiology; 2011;79:2597--607. Available from:
\url{https://doi.org/10.1128/iai.05118-11}

\hypertarget{ref-Ohkusa2003}{}
34. Ohkusa T. Induction of experimental ulcerative colitis by
fusobacterium varium isolated from colonic mucosa of patients with
ulcerative colitis. Gut {[}Internet{]}. BMJ; 2003;52:79--83. Available
from: \url{https://doi.org/10.1136/gut.52.1.79}

\hypertarget{ref-Ohkusa2002}{}
35. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
Fusobacterium varium localized in the colonic mucosa of patients with
ulcerative colitis stimulates species-specific antibody. Journal of
Gastroenterology and Hepatology {[}Internet{]}. Wiley-Blackwell;
2002;17:849--53. Available from:
\url{https://doi.org/10.1046/j.1440-1746.2002.02834.x}

@@

\subsubsection{Figures}\label{figures}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../submission/fig1.jpg}
\caption{Fig 1}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../submission/figure_2.pdf}
\caption{Fig 2}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../submission/figure_3.pdf}
\caption{Fig 3}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../submission/figure_4.pdf}
\caption{Fig 4}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../submission/figure_5.pdf}
\caption{Fig 5}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../submission/figure_6.pdf}
\caption{Fig 6}
\end{figure}

@@

Location and relative abundance of cancer-associated OTUs. Relative
abundance was calculated and plotted by sample site for each OTU of
interest: (A) \emph{Fusobacterium nucleatum subsp. animalis} (B)
\emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}
\includegraphics{../submission/figure_S1.pdf}

>>>>

The human colon is an ecosystem comprised of numerous microenvironments
that select for different microbiota. Concentrations of oxygen, water,
and anti-microbial peptides change along the gut axis and influence
which microbiota reside in each location. Microenvironments differ not
only longitudinally along the colon, but also radially from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. The identity of these
specific microbiota and communities are important for understanding the
etiology of complex diseases such as Colorectal Cancer (CRC) and
Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or
accelerated by perturbations of the structure of the gut microbiota
(1--3). The manifestations of these diseases are known to vary based
upon the location in which they occur. For instance, CRC that arises in
the distal (left) colon are of hindgut origin and tend to have large
chromosomal alterations indicative of chromosomal instability (1). In
contrast, CRC arising in the proximal (right) colon are of midgut origin
and tend to be sessile and microsatellite instable (MSI with BRAF and
KRAS mutations) (1). In addition to the environmental gradients within
the colon, the distal and proximal sides of the colon differ in the
amount of inflammation present and the genomic instability of
precancerous cells, respectively (1,4,5). In IBD patients, disease
occurring in the distal colon extending proximally is usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur
anywhere along the GI tract, most commonly in the ileum and the cecum
(2). UC presents as continuous disease with only mucosal involvement,
where as CD has skip lesions and full thickness involvement that may
cause abscesses, strictures and fistulas (2). Thus, given the varied
physiology of the proximal-distal axis of the colon and known
differences in disease patterns at these sites, symbiotic microbiota and
their metabolites likely vary as well, and may influence the
heterogeneous disease prognoses of IBD and CRC. Because CRC can be a
long-term complication of IBD, the distribution of microbiota is
important to understanding the pathophysiology of both diseases.

@@

The few existing profiles of the microbial spatial variation of the
colon have been limited by sample collection methods. The majority of
human gut microbiome studies have been performed on whole shed feces or
on samples collected during colonoscopy or surgery (5). While invasive
methods allow investigators to acquire samples from inside the human
colon, typically these procedures are preceded by the use of bowel
preparation methods such as the consumption of laxatives to cleanse the
bowel. Bowel preparation is essential for detecting cancerous or
precancerous lesions in the colon, but complicates microbiome profiling
as the chemicals strip the bowel of contents and disrupt the mucosal
layer (13,14). As such, what little information we do have about the
spatial distribution of the microbiota in the proximal and distal colon
is confounded by the bowel preparation procedure.

@@

The procedures in this study and consent were approved by the
Institutional Review Board at the University of Michigan Health System
with protocol number HUM00082721. Subjects were recruited using the
online recruitment platform and were pre-screened prior to enrollment in
the study. Exclusion criteria included: use of asprin or NSAIDs within 7
days, use of antibiotics within 3 months, current use of anticoagulants,
known allergies to Fentanyl, Versed and Benadryl, prior history of colon
disease, diabetes, abdominal surgery, respiratory, liver, kidney or
brain impairments, undergoing current chemotherapy or radiation
treatment and subjects that were pregnant or trying to conceive. 20
subjects that met the criteria were selected and provided signed
informed consent prior to the procedure. There were 13 female and 7 male
subjects ranging in age from 25 to 64. 18 of the 20 subjects had not
used antibiotics within a year prior to the collection date and 2 had
not used antibiotics within 6 months. None of the subjects had medical
conditions requiring frequent or extended antibiotic use.

@@

At a baseline visit, subjects gave consent and were given a home
collection stool kit (Zymo). One to seven days prior to the scheduled
colonoscopy, subjects collected whole stool at home and shipped the
samples to a research coordinator on ice. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On the procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Subjects were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and jumbo
biopsy forceps used to collect the luminal contents. Two luminal samples
were collected and the contents immediately deposited into RNAlater
(Fisher) and flash-frozen in liquid nitrogen. The forceps were withdrawn
and new biopsy forceps were used to collect mucosal biopsies on sections
of the colon that were pink and free of stool matter. Three mucosal
biopsies were collected and flash-frozen in RNAlater. These samples
comprised the distal colon samples. The sigmoidoscope was then withdrawn
and a pediatric colonoscope was inserted to reach the proximal colon.
The proximal samples were taken from the ascending colon proximal to the
hepatic flexure at 75-120cm depending on the subject. Samples were then
collected in the same manner as was done in the distal colon and the
colonoscope withdrawn. All samples were stored at -80\(^\circ\)C.

@@

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was normalized to equal concentrations across
all samples and used as template for amplification of the V4 region of
the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as
previously described (15). Sequences were curated using the mothur
software as described previously (16). The sequences were assigned a
taxonomic classification using a naive Bayesian classifier trained using
a 16S rRNA gene training set from the Ribosomal Database Project (RDP)
(17) and clustered into operational taxonomic units (OTUs) based on a
97\% similarity cutoff. Sequencing and analysis of a mock community
revealed the error rate to be 0.018\%. Samples were rarefied to 4231
sequences per sample in order to reduce the effects of uneven sampling
bias.

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
1.39.5 (16). \(\theta\)\textsubscript{YC} distances were calculated to
determine the dissimilarity between two samples. Random Forest
classification models were built using the AUCRF R package using a
leave-one-subject out approach (18). The Random Forest models were built
using the full, non-rarefied, dataset as input. For each model the data
were split into a 19-subject training set and a 1-subject test set. The
model was built and cross-validated using 10-fold k cross-validation
(AUCRFcv) on the training set to estimate the prediction error of the
model. The resultant model was then used to predict the outcome the
left-out subject. This process was repeated iteratively for all subjects
and results plotted as Reciever Operator Characteristic curves using the
pROC R package (19). Resultant models were used to identify the OTUs
that were most important for classifying each location. Species-level
information for sequences of interest was obtained by aligning the
sequences to the GenBank nucleotide databse using blastn. The species
name was only used if the identity score was \(\ge\) 99\% over the
full-length of the contig and matched a single reference.

@@

When examining the relative abundance of the dominant phyla at each site
(i.e. \emph{Bacteriodes} and \emph{Firmicutes}), there was a wide amount
of variation. This likely reflects not only the variability between
human subjects, caused by differences in age, sex, and diet, but may
also reflect spatial patchiness in the gut microbiome within a subject.
Patchiness refers to inconsistent distribution of microbial populations
due to fluctuations in local resources (26). One study noted that the
bacteria recoverable from the same mucosal sample location can be vastly
different when the samples are taken just 1 cm away from each other
(27). Similar patchiness was also observed in luminal contents and fecal
samples themselves; there was separation of different interacting
microbes along the length of a stool sample, for instance (28). A third
study that sampled six mucosal sites along the colon observed such
patchiness in two of the three study subjects (21). While our subjects
were not sampled frequently enough to draw specific conclusions about
patchiness along the unprepped colon, we did observe some specific
differences in mucosal versus luminal samples at the phylum level. The
mucosal samples harbored more \emph{Proteobacteria}, consistent with
previous studies comparing mucosal swabs to luminal content in humans
(4). However, we must still consider that the results from phyla
analysis may have been impacted by inter-subject patchiness.

@@

In addition to identifying features that are specific to each side of
the gut, the ability of the Random Forest to classify samples can serve
as a proxy for similarity. That is, a higher AUC value indicates the
samples are more efficiently classfied (and thus more different) than a
model with a lower AUC value. For instance, the model separating the
proximal and distal mucosa had an AUC of 0.850 whereas the model for
classifying the proximal and distal lumen had a much lower AUC of 0.580.
Further, the latter model required 44 OTUs to best separate the samples
whereas the models separating the mucosa only needed 10 OTUs. The much
lower AUC and need for a high number of features compared to other
models suggest these locations are the most similar of the comparisons
tested. We speculate that the model was less effective at classifying
the proximal and distal luminal contents because the mucosal
microenvironments have more variable selective pressure along the colon
than the luminal microenvironments.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 8 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans (9) and have oncogenic
properties in cell culture and in mice (29). Though the bacteria are
known to co-localize on CRC tumors, in our study \emph{F. nucleatum} was
found on both sides of the colon while \emph{P. asacharolytica} was only
detected in the distal mucosa. Not much is known about the distribution
of \emph{P. asacharolytica} along the healthy colon, but given its
anaerobic lifestyle and asacharolytic metabolism, it is perhaps not
surprising that our study detected the bacteria primarily in the
less-oygen-rich and protein-rich distal mucosa (4). In studies examining
bacteria on colorectal cancer tumors, \emph{F. nucleatum} was more
commonly detected on proximal-sided tumors, and distribution of \emph{F.
nucleatum} decreased along the colon to rectum (30). Of the 8 (40\%)
individuals positive for \emph{F. nucleatum} in our study, the bacterium
was spread across the proximal mucosa, distal lumen and distal mucosa.
The \emph{Fusobacterium} species \emph{nucleatum} and \emph{varium} have
been commonly isolated from mucosal biopsies of patients with IBD and UC
(25,31). In our study, \emph{F. varium} was only detected in three
subjects and two of those samples were isolated form the proximal mucosa
(Supplementary Fig. S1B). \emph{F. varium} is most commonly isolated
from UC patient biopsies from the ileum or cecum (adjacent to the
proximal colon) (32), suggesting this species may exhibit preference for
the higher oxygen content of these gastrointestinal sites.

Spatial organization of \emph{Fusobacteria} and other bacterial species
into polymicrobial biofilms that can invade the gut mucosa have been
linked to CRC (33). The biofilms promote tumorigenesis by allowing
bacteria to grow near the epithelium, inducing inflammation,
genotoxicity and metabolic changes that favor tumor cell growth (33). In
one study of CRC biofilms and tumors, all of the proximal tumors
examined contained a polymicrobial biofilm on the tumor mucosa (7).
Further examination of these tumors identified \emph{Fusobacteria}
species as members of the biofilm. These results indicate that it is not
only the presence of the bacterium but the tumor community as a whole
that contributes to tumorigenesis (7). A `driver-passenger' model has
been proposed as a mechanism for biofilm assembly in the gut (34,35). In
this model, `driver' species such as \emph{Fusobacterium spp} and
\emph{Porphyromonas spp} exert tumorigenic effects locally and create a
niche for adherence of `passenger' species that comprise the rest of the
biofilm (34,35). Thus the the distribution of these disease-associated
microbes in healthy patients is of interest as their presence can be
predictive of disease prior to the onset of symptoms (9). A better
understanding of the early microbial changes in the gut microboiome is
essential for elucidating a mechanism for development of CRC or IBD
subtypes in the proximal or distal colon.

@@

By revealing specific differences in microbial populations at each
location in the gut via sampling an unprepared bowel, we can begin to
form hypotheses about how specific host-microbe interactions can affect
disease progression of proximal and distal CRC and IBD subtypes. Future
investigation of these samples using metagenomics and
metatranscriptomics would illuminate the microbial activities in these
gut microenvironments. Further, combining this approach with a more
comprehensive sampling strategy along the unprepped colon could enhance
microbiome-based screening and treatment modalities for colon disease.

@@

\hypertarget{ref-Rougharden1977}{}
26. Rougharden JD. Patchiness in the spatial distribution of a
population caused by stochastic fluctuations in resources. Oikos
{[}Internet{]}. JSTOR; 1977;29:52. Available from:
\url{https://doi.org/10.2307/3543292}

\hypertarget{ref-Hong2011}{}
27. Hong P-Y, Croix JA, Greenberg E, Gaskins HR, Mackie RI.
Pyrosequencing-based analysis of the mucosal microbiota in healthy
individuals reveals ubiquitous bacterial groups and micro-heterogeneity.
Ahmed N, editor. PLoS ONE {[}Internet{]}. Public Library of Science
(PLoS); 2011;6:e25042. Available from:
\url{https://doi.org/10.1371/journal.pone.0025042}

\hypertarget{ref-Stearns2011}{}
28. Stearns JC, Lynch MDJ, Senadheera DB, Tenenbaum HC, Goldberg MB,
Cvitkovitch DG, et al. Bacterial biogeography of the human digestive
tract. Scientific Reports {[}Internet{]}. Springer Nature; 2011;1.
Available from: \url{https://doi.org/10.1038/srep00170}

\hypertarget{ref-Sears2014}{}
29. Sears CL, Garrett WS. Microbes, microbiota, and colon cancer. Cell
Host \& Microbe {[}Internet{]}. Elsevier BV; 2014;15:317--28. Available
from: \url{https://doi.org/10.1016/j.chom.2014.02.007}

\hypertarget{ref-Mima2016}{}
30. Mima K, Cao Y, Chan AT, Qian ZR, Nowak JA, Masugi Y, et al.
Fusobacterium nucleatum in colorectal carcinoma tissue according to
tumor location. Clinical and Translational Gastroenterology
{[}Internet{]}. Springer Nature; 2016;7:e200. Available from:
\url{https://doi.org/10.1038/ctg.2016.53}

\hypertarget{ref-Ohkusa2003}{}
31. Ohkusa T. Induction of experimental ulcerative colitis by
fusobacterium varium isolated from colonic mucosa of patients with
ulcerative colitis. Gut {[}Internet{]}. BMJ; 2003;52:79--83. Available
from: \url{https://doi.org/10.1136/gut.52.1.79}

\hypertarget{ref-Ohkusa2002}{}
32. Ohkusa T, Sato N, Ogihara T, Morita K, Ogawa M, Okayasu I.
Fusobacterium varium localized in the colonic mucosa of patients with
ulcerative colitis stimulates species-specific antibody. Journal of
Gastroenterology and Hepatology {[}Internet{]}. Wiley-Blackwell;
2002;17:849--53. Available from:
\url{https://doi.org/10.1046/j.1440-1746.2002.02834.x}

\hypertarget{ref-Li2017}{}
33. Li S, Konstantinov SR, Smits R, Peppelenbosch MP. Bacterial biofilms
in colorectal cancer initiation and progression. Trends in Molecular
Medicine {[}Internet{]}. Elsevier BV; 2017;23:18--30. Available from:
\url{https://doi.org/10.1016/j.molmed.2016.11.004}

\hypertarget{ref-Tjalsma2012}{}
34. Tjalsma H, Boleij A, Marchesi JR, Dutilh BE. A bacterial
driverpassenger model for colorectal cancer: Beyond the usual suspects.
Nature Reviews Microbiology {[}Internet{]}. Springer Nature;
2012;10:575--82. Available from:
\url{https://doi.org/10.1038/nrmicro2819}

\hypertarget{ref-Flynn2016}{}
35. Flynn KJ, Baxter NT, Schloss PD. Metabolic and community synergy of
oral bacteria in colorectal cancer. McMahon K, editor. mSphere
{[}Internet{]}. American Society for Microbiology; 2016;1:e00102--16.
Available from: \url{https://doi.org/10.1128/msphere.00102-16}

@@

\subsubsection{Figure legends}\label{figure-legends}

@@

@@

@@

@@

@@

@@

@@

Location and relative abundance of cancer-associated OTUs. Relative
abundance was calculated and plotted by sample site for each OTU of
interest: (A) \emph{Fusobacterium nucleatum subsp. animalis} (B)
\emph{Fusobacterium varium} and (C) \emph{Porphyromonas asacharolytica}
